### 2023-04-24 response-to-audit-report.md

# Response to Audit Report

**TODO:**

- [x] Add prefatory text and executive summary.
- [x] Add responses and mitigation commits to Isabelle-related findings.
- [x] Add responses and mitigation commits to Haskell-related findings.
- [ ] Fix display of equations or replace ones that cannot be fixed with ellipses ("...").
- [ ] Replace broken links and cross references, or use ellipses ("...").
- [ ] Verify the section numbering is correct.
- [ ] Add table of contents.
- [ ] Proofread.

@@

## Acknowledgement

We thank the High Assurance Software Group at Tweag for their diligent and expert analyses of the Marlowe specifications, proofs, implementations, and tests and their clear and thorough communication of their findings.

## Overview of Isabelle Changes

The audit report's main and high-severity concerns for the Isabelle Proofs were:

- 2.1.6 Missing description of Merkleization
- 2.1.4 Inaccurate formulation of Money preservation

We didn't mention the merkleization in the specification document as we thought of it as an implementation detail. This does lessen the applicability of the Isabelle proofs to the different implementations (marlowe-cardano, purescript-marlowe, etc). To check that an implementation is compliant to the specification we use the `spec-test` tool, which uses property based tests to check that a custom implementation behaves equally to the Isabelle implementation. Without the proper formalization, the tool can only test for the non-merkleized cases. In practice this gives us confidence that the merkleized interpreter works the same for non-merkleized cases. To test the correctness of merkleized cases, the marlowe-cardano repository has specific tests. The Merkleization formalization will be introduced in a future version of the specification, the initial formulation started in [PR-171](https://github.com/input-output-hk/marlowe/pull/171).

The Money preservation theory was only taking into account the amount of assets and not the type. This meant that the proof wasn't considering a case were a contract could receive `20 ada`, `15 djed` and pay back `20 djed` and `15 ada`. We added a new MultiAssets type and refactored the Isabelle code (without modifying the interpreter) to prove that the Assets are preserved.

Other medium and low priorities findings regarding documentation and legibility of the specification were also addressed.

@@

Although this `Eq` instance is not used by the Marlowe validators (so it does not have implications for validator security), the equality test has been fixed because it is a liability for off-chain code and for potential future versions of validators that might perform such an equality test. Commit [84d65a70](https://github.com/input-output-hk/marlowe-cardano/commit/84d65a70c29320fab501ad759bbc71855d1b63fc) fixes this and commit [04805a39](https://github.com/input-output-hk/marlowe-cardano/commit/04805a393c0815fe0f2a1a6d7f76d9867ffe0c14) hardens all of the other `Eq` instances to prevent similar problems re-arising if sum types are augmented in the future.

### 2.1.4 Inaccurate formulation of Money preservation *(Severity: High)*

> **File `specification-v3-rc1.pdf`, Section `3.1 Money preservation`, page *29***
>
> As the property stands, it is permitted to make deposits in one currency and return payments in a different currency. As long as the sums of the amounts match, the equality is satisfied. Yet it is unlikely that the participants of the contract would agree that money has been preserved.
>
> Money preservation is a property stated with an equality. The left hand side is the sum of the deposits done by a list of transactions. The right hand side of the equality is the sum of all the payments done in the same list of transactions. Each sum, in turn, is represented as a single integer which aggregates the amounts of the various payments and deposits, irrespective of what currencies correspond to these amounts.

To address this, a new `MultiAssets.thy` theory was created. This theory defines a new type that ties together the number of tokens and the type of token, and it implements some type classes to be able to work with normal addition and substraction. Also the MoneyPreservation was refactored into AssetsPreservation, addressing issue 2.1.4 among others.

These changes can be found on [PR-161](https://github.com/input-output-hk/marlowe/pull/161)

@@

### 2.1.5 Insufficient documentation of Money preservation *(Severity: Medium)*

> **File `specification-v3-rc1.pdf`, Section `3.1 Money preservation`, page *29***
>
> Money preservation is formulated in terms of functions that are not discussed in the specification. It is necessary to explain the meaning of these functions in sufficient detail so readers can understand the property.

The theory was completly refactored in [PR-161](https://github.com/input-output-hk/marlowe/pull/161), adding the necessary documentation.

@@

### 2.1.6 Missing description of Merkleization *(Severity: High)*

> **File `specification-v3-rc1.pdf`, `Merkleization`**
>
> There is no property about merkleization, but merkleization is implemented in the Cardano integration.
>
> Some relevant properties could be:
>
> 1.  The merkleized contract produces the same payments as the analogous regular contract.
> 2.  If a merkleized case input is applied successfully, it implies that the contract hash in the input corresponds to the continuation of the contract.
> 3.  Merkleizing and unmerkleizing a contract gives back the original contract.

We didn't mention the merkleization in the specification document as we thought of it as an implementation detail. This does lessen the applicability of the Isabelle proofs to the different implementations (marlowe-cardano, purescript-marlowe, etc). To check that an implementation is compliant to the specification we use the `spec-test` tool, which uses property based tests to check that a custom implementation behaves equally to the Isabelle implementation. Without the proper formalization, the tool can only test for the non-merkleized cases. In practice this gives us confidence that the merkleized interpreter works the same for non-merkleized cases. To test the correctness of merkleized cases, the marlowe-cardano repository has specific tests.

The Merkleization formalization will be introduced in a future version of the specification, the initial formulation started in [PR-171](https://github.com/input-output-hk/marlowe/pull/171).

@@

Again as previously, the relevant commits are [0a890845](https://github.com/input-output-hk/marlowe-cardano/commit/0a890845c44ccfe4df97e765e9b9eb743ce7d580), [8855feae](https://github.com/input-output-hk/marlowe-cardano/commit/8855feae473882b82643db792327423152e45b5c), [26f024e8](https://github.com/input-output-hk/marlowe-cardano/commit/26f024e80cea0dd4ebf8bff0f1a10b97aea87894), and [201c5df9](https://github.com/input-output-hk/marlowe-cardano/commit/201c5df9de29b3b142be5ff7d0a9b56d1b378204) for specification and implementation and [d514bcd0](https://github.com/input-output-hk/marlowe-cardano/commit/d514bcd075732b07e1c6a73e2e3c68afa01acf2c), [7f562545](https://github.com/input-output-hk/marlowe-cardano/commit/7f562545d013dd17472b692b550ffdc7f8a383f3), and [ebab31d9](https://github.com/input-output-hk/marlowe-cardano/commit/ebab31d9ba6dd322d374b34cad3995050a0e476e) for property-based tests.

### 2.1.11 Partial functions used outside their domain *(Severity: Medium)*

> **File `MoneyPreservation.thy`, various functions ``**
>
> `moneyInRefundOneResult`, `moneyInApplyResult`, `moneyInApplyAllResult`, `moneyInTransactionOutput`, and `moneyInPlayTraceResult` have strange meanings when the result is an error. Arguably, on error there is no money to retrieve, so the return type should be `(Token \times int) option` instead.
>
> Some lemmas rely on this behavior to have equalities hold even in cases of errors, but the cost is that the meaning is so surprising that the reader may be confused by it. It would be more reliable to have explicit and weaker lemmas that assert equalities only when there are no errors.

This module was refactored in [PR-161](https://github.com/input-output-hk/marlowe/pull/161) and these functions were no longer needed.

@@

## 2.2 Marlowe specification

### 2.2.1 Lack of explanation regarding changing choices *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.1.4 Choices`, page *10***
>
> Choices can only be changed when evaluating `When` statements. This is something only evident after looking at the implementation of `computeTransaction`. It needs to be discussed when first introducing choices and the `When` contract.

We improved documentation regarding choices in [PR-168](https://github.com/input-output-hk/marlowe/pull/168) and [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

@@

### 2.2.2 Undefined reference *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.1.7 Contracts`, page *13***
>
> There is an undefined reference.

We fixed this in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

@@

### 2.2.3 Lack of explanation for necessity of `Environment` type *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.1.8 State and Environment`, page *14***
>
> An `Environment` type is introduced, but it is unclear why it is needed as it is defined as a synonym for time intervals.

We added more documentation regarding the environment in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

@@

### 2.2.4 Unclear meaning of execution environment *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.1.8 State and Environment`, page *14***
>
> The meaning of the execution environment of the transaction is unclear. This is due to the concept of *transaction* being assumed by the specification and never formally introduced.
>
> The specification reads
>
> > The execution environment of a Marlowe contract simply consists of the (inclusive) time interval within which the transaction is occurring.
>
> One has to infer that evaluating a Marlowe contract is undefined if it does not happen within a transaction, as otherwise the description of the execution environment would not make sense. It would be necessary to establish more explicitly the relationship between the contract evaluation and the notion of transaction.

We added more documentation regarding the execution environment in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

@@

### 2.2.5 Unexplained interval data types *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.1.8 State and Environment`, page *14***
>
> The meaning of the data types `IntervalError` and `IntervalResult` needs to be explained.

These types were properly documented in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

@@

### 2.2.6 Incomplete explanation for `TransactionOutput` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.1 Compute Transaction`, page *15***
>
> The meaning of the data type `TransactionOutput` needs to be explained. More generally, the meaning of the return types of most functions has to be explained. Currently, the meaning can only be inferred from looking at how the types are used, which makes it harder to identify if they are used as intended.
>
> The purpose of these types needs to be made explicit so it can be checked if the code is doing what is intended.

This type was properly documented in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

@@

### 2.2.7 Code snippets switch languages *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.1 Compute Transaction`, page *15***
>
> The specification changes from using Isabelle to using Haskell henceforth. Making the reader aware of the criteria for the language change would help maintaining the document.

The cause of this problem was that initially we had the Isabelle code separated from the documentation generation and we coulnd't find a way to display the Isabelle code, so we showed the exported Haskell instead. We agree that it is confusing so we refactored to a literate programming style in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

@@

### 2.2.8 Repeated definition of `IntervalResult` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Sections `2.1.8 State and Environment, 2.2.2 Fix Interval`, pages *14, 16***
>
> The `IntervalResult` type is defined twice in the specification. One should be removed.

We removed the extra definition in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)

@@

### 2.2.9 Poorly named variable `newAccount` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.6 Reduce Contract Step`, page *19***
>
> In the implementation of the function `reduceContractStep`, the variable `newAccount` should be named `newAccounts`.

We renamed as suggested in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

@@

We renamed as suggested in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

### 2.2.11 Inaccurate specification of `giveMoney` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.9 Utilities`, page *22***
>
> It says
>
> > The *giveMoney* function transfers funds internally between accounts.
>
> which is not accurate. It should say instead
>
> > The *giveMoney* function deposits funds to an internal account.
>
> This function is confusing in that it takes the account identifier of the paying account which is not used for anything other than filling a field in the returned value.

This function is confusing, we are likely to remove it in a future release.

@@

### 2.2.12 Redundant evaluation in `addMoneyToAccount` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.9 Utilities`, page *22***
>
> `addMoneyToAccount` is redundantly evaluating `money <= 0` when invoking `updateMoneyInAccount`. The else branch could be replaced instead with `insert (accId, token) money accountsV`.

We disagree on this observation. The function `updateMoneyInAccount` checks if its money parameter is non positive to always have accounts with positive balance. The
`addMoneyToAccount` function calls `updateMoneyInAccount` with `balance + money` and checks that `money` is positive. This makes sure that `addMoneyToAccount` always increases the assets.

@@

### 2.2.13 Redundant statement regarding addition *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.10 Evaluate Value`, page *24***
>
> It says that addition is associative and commutative. This is true but it is already implied by the equation preceding the statement. Maybe change to
>
> > Note that addition is associative and commutative.
>
> or remove the redundant statement.

The statement was removed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)

@@

### 2.2.14 Missing implementation for negation case of `evalValue` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.10 Evaluate Value`, page *24***
>
> Negation for `evalValue` does not show the implementation, just one lemma about `NegValue`, which is inconsistent with how other operations are presented.

The way we document evalValue was modified in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

@@

### 2.2.15 Missing parentheses in `div` specification *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.10 Evaluate Value`, page *25***
>
> On page 25 formula $$c \neq 0 \Rightarrow c \mathbin{∗} a \mathbin{\mathrm{div}} (c \mathbin{∗} b) = a \mathbin{\mathrm{div}} b$$ needs additional parentheses around the term $c \mathbin{∗} a$, otherwise it can be parsed as $$c \neq 0 \Rightarrow c \mathbin{∗} (a \mathbin{\mathrm{div}} (c \mathbin{∗} b)) = a \mathbin{\mathrm{div}} b$$ which does not hold (Counter-example: $c=2, a=3, b=2$). The lemma `divMultiply` in the file `Semantics.thy` does use extra parentheses around $c \mathbin{∗} a$.

This lemma was removed from the PDF specification and replaced with lemmas that uses `evalValue` instead.

@@

### 2.2.16 Unclear division explanation *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.10 Evaluate Value`, page *25***
>
> It says
>
> > Division is a special case because we only evaluate to natural numbers.
>
> The meaning of this statement needs to be further explained, since the arguments of `DivValue` could evaluate to negative numbers.

This was corrected to use `integer` instead of `natural` and further rephrased in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

@@

The way we document evalValue was modified in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

### 2.2.18 Missing `evalValue` lemmas in specification *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.10 Evaluate Value`, pages *23--26***
>
> Not all lemmas about `evalValue` are listed in the specification. The absent lemmas include `evalDoubleNegValue`, `evalMulValue`, `evalSubValue`, and all division lemmas.

This section was rewritten an only the Division lemmas are now present as they are the most relevant ones.

@@

This section was rewritten and the issue was fixed

### 2.2.20 Unexplained parameters of `playTrace` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `3 Marlowe Guarantees`, page *28***
>
> The parameters of the function `playTrace` need to be explained.

The function `playTrace` was documented in [PR-182](https://github.com/input-output-hk/marlowe/pull/182) in section 2.2.7.

@@

### 2.2.21 Type parameter discrepancy in `playTrace` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `3 Marlowe Guarantees`, page *28***
>
> The first parameter of `playTrace` in the specification is `int`, while it is `POSIXTime` in the code. Even though the latter is an alias for the former, it is beneficial to use the `POSIXTime` name both for consistency and readability.

This happened because we were using a type alias, which gets fully expanded when being referenced by the documentation.  [PR-182](https://github.com/input-output-hk/marlowe/pull/182) address this by making the whole theory in literal programming mode.

@@

### 2.2.22 Money preservation on failing transactions not specified *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `3.1 Money preservation`, page *29***
>
> Money preservation is expressed with an equality. This equality, however, only ensures money preservation for those lists of transactions that produce no error. In other words, there is no guarantee that money will be preserved for those lists of transactions that fail.
>
> This is not a concern in practice because the lists of transactions that fail to evaluate are not accepted in the blockchain. However, this should be made explicit in the explanation of the property.

This was taken into account in the new Asset preservation theory and a [paragraph](https://github.com/input-output-hk/marlowe/pull/161/files#diff-060b0c9f7bde4e9ebc4e7b9d6a2202af75213be8d44d7f4745e02054be5186deR52-R53) was added to explain the case.

@@

### 2.2.23 Complicated definition of `allAccountsPositive` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `3.3 Possitive Accounts`, page *30***
>
> The definition of\ `allAccountsPositive` is complicated and can be refactored as `all ((_, money) -> money > 0)`.

The proposed definition is simpler and it is likely to be implemented in the future.

@@

### 2.2.24 Discrepancy with Isabelle code for `allAccountsPositive` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Sections `3.3 Positive Accounts`, page *30***
>
> The `allAccountsPositive` function is defined differently in the specification and in the Isabelle code, although both definitions show the same behavior. These definitions need to be consolidated.

The definition in the specification is the exported haskell code of the Isabelle code. This is a limitation that Isabelle has when importing code from other theories. A possible solution is to change the Accounts positive theory to be literal programming.

@@

### 2.2.25 Misleading or incorrect formula for contract not holding funds *(Severity: Low)*
